\chapter{Results and Analysis}
\label{ch:results}

This chapter presents the experimental results of the policy formalization system, including the LLM model comparison, rule extraction accuracy, and formal verification outcomes.

\section{LLM Model Comparison}

\subsection{Experimental Setup}

We evaluated five large language models for their ability to identify policy rules in academic documents. The models were selected based on published benchmarks and peer-reviewed research (Table \ref{tab:models-tested}).

\begin{table}[h]
\centering
\caption{LLM Models Evaluated for Policy Rule Identification}
\label{tab:models-tested}
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model} & \textbf{Parameters} & \textbf{Organization} & \textbf{Reference} \\
\hline
Llama 3.2 & 3B & Meta AI & Meta (2024) \\
Phi-3 & 3.8B & Microsoft Research & Abdin et al. (2024) \\
Mistral & 7B & Mistral AI & Jiang et al. (2023) \\
Mixtral & 47B (13B active) & Mistral AI & Jiang et al. (2024) \\
Llama 3.1 & 70B & Meta AI & Meta (2024) \\
\hline
\end{tabular}
\end{table}

All models used identical prompts and were run with temperature 0.1 for deterministic outputs. The evaluation was conducted on 97 policy rules extracted from AIT policy documents.

\subsection{Classification Performance}

Table \ref{tab:model-performance} presents the classification statistics for each model.

\begin{table}[h]
\centering
\caption{LLM Model Classification Performance (n=97 rules)}
\label{tab:model-performance}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Model} & \textbf{Rules} & \textbf{Not Rules} & \textbf{Errors} & \textbf{Rule Rate} & \textbf{Confidence} \\
\hline
\textbf{Mistral} & \textbf{96} & \textbf{1} & \textbf{0} & \textbf{99.0\%} & \textbf{0.98} \\
Mixtral & 95 & 0 & 2 & 97.9\% & 0.91 \\
Llama 3.2 & 93 & 0 & 4 & 95.9\% & 0.88 \\
Phi-3 & 92 & 1 & 4 & 94.8\% & 0.91 \\
Llama 3.1 & 90 & 5 & 2 & 92.8\% & 0.89 \\
\hline
\end{tabular}
\end{table}

\textbf{Key Finding:} Mistral 7B achieved the highest performance with 99.0\% rule detection rate and 0.98 average confidence, outperforming larger models including Llama 3.1 (70B) which achieved only 92.8\%. This finding aligns with recent research indicating that model size does not always correlate with classification performance.

\subsection{Inter-Model Agreement}

Table \ref{tab:model-agreement} shows pairwise agreement rates between models.

\begin{table}[h]
\centering
\caption{Inter-Model Agreement Rates}
\label{tab:model-agreement}
\begin{tabular}{|l|c|l|}
\hline
\textbf{Model Pair} & \textbf{Agreement} & \textbf{Interpretation} \\
\hline
Mistral vs Mixtral & 96.9\% & Excellent \\
Phi-3 vs Mistral & 95.9\% & Excellent \\
Llama 3.2 vs Mistral & 94.8\% & Excellent \\
Mistral vs Llama 3.1 & 93.8\% & Excellent \\
Llama 3.2 vs Llama 3.1 & 88.7\% & Good \\
\hline
\end{tabular}
\end{table}

All model pairs demonstrated agreement rates above 88\%, indicating reliable classification across models. The highest agreement (96.9\%) was observed between Mistral 7B and Mixtral 47B.

\subsection{Model Selection Justification}

Based on the comparison results, \textbf{Mistral 7B} was selected as the primary model for this research for the following reasons:

\begin{enumerate}
    \item \textbf{Highest accuracy}: 99.0\% rule detection rate
    \item \textbf{Highest confidence}: 0.98 average confidence score
    \item \textbf{Zero parsing errors}: Only model with no JSON parsing failures
    \item \textbf{Computational efficiency}: 7B parameters vs 70B for comparable accuracy
    \item \textbf{Cross-validation}: 96.9\% agreement with larger Mixtral model
\end{enumerate}

\section{Rule Extraction Analysis}

\subsection{Corpus Statistics}

The policy corpus consists of 492 potential rules extracted from 5 AIT policy documents. Table \ref{tab:rule-distribution} shows the distribution of rule types as classified by Mistral.

\begin{table}[h]
\centering
\caption{Distribution of Rule Types in Policy Corpus}
\label{tab:rule-distribution}
\begin{tabular}{|l|c|c|}
\hline
\textbf{Rule Type} & \textbf{Count} & \textbf{Percentage} \\
\hline
Obligation & 45 & 46\% \\
Permission & 28 & 29\% \\
Prohibition & 18 & 19\% \\
Recommendation & 6 & 6\% \\
\hline
\textbf{Total} & \textbf{97} & \textbf{100\%} \\
\hline
\end{tabular}
\end{table}

The predominance of obligations (46\%) is consistent with the prescriptive nature of academic policy documents.

\subsection{Deontic Marker Analysis}

Table \ref{tab:deontic-markers} presents the deontic markers identified in the corpus.

\begin{table}[h]
\centering
\caption{Deontic Markers Identified in Policy Rules}
\label{tab:deontic-markers}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Marker} & \textbf{Type} & \textbf{Example} \\
\hline
must & Obligation & ``students \textbf{must} pay fees'' \\
shall & Obligation & ``accounts \textbf{shall} be reviewed'' \\
may & Permission & ``students \textbf{may} opt to...'' \\
cannot & Prohibition & ``\textbf{cannot} enroll in any course'' \\
should & Recommendation & ``\textbf{should} seek explanation'' \\
required & Obligation & ``\textbf{required} to vacate rooms'' \\
\hline
\end{tabular}
\end{table}

\section{Formal Verification Results}

\subsection{FOL Formalization}

Policy rules were formalized into First-Order Logic (FOL) using deontic operators:

\begin{itemize}
    \item $O(\phi)$: Obligation - $\phi$ must happen
    \item $P(\phi)$: Permission - $\phi$ may happen
    \item $F(\phi)$: Prohibition - $\phi$ is forbidden
\end{itemize}

\subsubsection{Example Formalization}

\textbf{Policy Rule:} ``Self-support enrolled students must pay tuition fees before registration.''

\textbf{FOL Representation:}
\[
\forall s. \text{Student}(s) \land \text{IsSelfSupport}(s) \land \text{IsEnrolled}(s) \rightarrow O(\text{PayTuition}(s))
\]

\textbf{SHACL Translation:}
\begin{lstlisting}[language=Turtle]
ait:TuitionComplianceShape a sh:NodeShape ;
    sh:targetClass ait:Student ;
    sh:sparql [
        sh:select """
            SELECT $this WHERE {
                $this ait:hasStudentType "Self-Support" .
                $this ait:hasEnrollmentStatus "Enrolled" .
                $this ait:hasObligation ?obl .
                ?obl ait:hasObligationType "TuitionFee" .
                ?obl ait:isPaid false .
            }
        """ ;
        sh:message "Self-Support enrolled students must pay tuition" ;
    ] .
\end{lstlisting}

\section{Summary}

The experimental results demonstrate:

\begin{enumerate}
    \item \textbf{Model Selection}: Mistral 7B is optimal for policy rule classification with 99\% accuracy and 0.98 confidence
    \item \textbf{Model Size}: Larger models do not guarantee better classification performance; Mistral 7B outperformed Llama 3.1 70B
    \item \textbf{Inter-Model Reliability}: All model pairs show $>$88\% agreement, validating classification reliability
    \item \textbf{Rule Distribution}: Academic policies are predominantly obligatory (46\%) with clear deontic markers
    \item \textbf{Formalization}: Policy rules can be systematically converted to FOL and SHACL for automated verification
\end{enumerate}
